---
title: "Exercises 6"
author: "Jan Hohenheim"
output:
  pdf_document: default
  html_notebook: default
---

## Setup

```{r loading}
rm(list = ls())
library(ggplot2)
library(ggthemes)
library(dplyr)
library(purrr)
library(cowplot)
library(latex2exp)
library(glue)

theme <- 
  theme_solarized() #+ 
  #theme(text = element_text(family = "Helvetica Neue"))
```

## Problem 19 (Binomial distribution)
Suppose that among n = 95 Swiss males, eight are red-green colour blind. We are interested in estimating the proportion p of people suffering from such disease among the male population.

### a) Is a binomial model suitable for this problem?

Yes, using $X \sim Bin(n_{\rm{tot}}, \frac{8}{95})$, where $n_{\rm{tot}}$ is the male population.

### b) Calculate the maximum likelihood estimate (ML) $p_{\rm{ML}}$.

```{r p19b}
n <- 95
x <- 8
p_ml <- x/n
glue("P_ML: {p_ml}")
```

### d) Use the R functions `binom.test()` and `prop.test()` to compute two-sided 95%-confidence intervals for the exact and for the approximate proportion. Compare the results.

```{r p19d}
binom.test(x, n)
prop.test(x, n)
```
The normal approximation is a little bit higher than the exact test.

### e) What is the meaning of the p-value?

The p-value tells us the likelihood of a data at least as extreme as the one presented existing under H0. In this case, H0 postulates that p is equal to 0.5. We can see, that this is highly unlikely.

### f) Compute the Wilson 95%-confidence interval and compare it to the confidence intervals from d).

```{r p19f}
wilson.ci <- function(x, n) {
  q <- 1.96
  mid <- (x + q^2 / 2) / (n + q^2)
  se <- sqrt(n) / (n + q^2) * sqrt( x / n * (1 - x / n) + q^2 / (4 * n))
  cbind(pmax(0, mid - q * se), pmin(1, mid + q*se))
}
wilson.ci(x, n)
```
The Wilson CI algorithm produces the most narrow CI.

## Problem 20 (A simple clinical trial)

A clinical trial is performed to compare two treatments, A and B, that are intended to treat a skin disease named psoriasis. The outcome shown in the following table is whether the patient’s skin cleared within 16 weeks of the start of treatment.  
Use $\alpha = 0.05$ throughout this problem.

### a) Compute for each of the two treatments a Wald type and a Wilson confidence interval for the proportion of patients whose skin cleared
```{r p20a}
wald.ci <- function(x, n) {
  q <- 1.96
  mid <- x / n
  se <- sqrt(x * (n - x) / n^3)
  cbind(max(0, mid - q * se), min(1, mid + q * se))
}

treatment_a <- data.frame(cleared = 9,
                          not_cleared = 18)
treatment_b <- data.frame(cleared = 5,
                          not_cleared = 22)
print("95% CIs for treatment A:")
print("Wald:")
wald.ci(treatment_a$cleared, sum(treatment_a))
print("Wilson:")
wilson.ci(treatment_a$cleared, sum(treatment_a))

print("95% CIs for treatment B:")
print("Wald:")
wald.ci(treatment_b$cleared, sum(treatment_b))
print("Wilson:")
wilson.ci(treatment_b$cleared, sum(treatment_b))
```
Treatment A seems to be way better

### b) Test whether the risk difference is significantly different to zero (i.e., RD = 0). Use both an exact (test of proportions) and an approximated approach (See 6.4.1 in the script)
```{r p20b}
tab <- matrix(c(9, 5, 18, 22), ncol = 2, byrow = TRUE)
colnames(tab) <- c('A','B')
rownames(tab) <- c('cleared','not_cleared')
tab

prop.test(tab, n = NULL)

n1 <- sum(treatment_a)
n2 <- sum(treatment_b)
p1 <- treatment_a$cleared / n1
p2 <- treatment_b$cleared / n2
mu <- p1 - p2
sigma <- sqrt(p1*(1 - p1)/n1 + p2*(1 - p2)/n2)
c(qnorm(0.025, mu, sigma), qnorm(0.975, mu, sigma))
```
The exact and the approximate approach both accept H0

### c) Compute CIs for both, relative risk (RR) and odds ratio (OR).

```{r p20c}
print("RR:")
rr <- p1 / p2
s <- sqrt(1/tab[1,1] + 1/tab[2,1] - 1/sum(tab[1,]) - 1/ sum(tab[2,]))
exp(log(rr) + qnorm(c(.025, .975))*s)

print("OR:")
or <- tab[1]*tab[4]/(tab[2]*tab[3])
s <- sqrt(sum(1/tab))
exp(log(or) + qnorm(c(.025, .975))*s)
```

### d) How would the point estimate of the odds ratio change if we considered the proportions of patients whose skin did not clear?

```{r p20d}
print("OR of not cleared skin:")
or <- (tab[2]/tab[1])/(tab[4]/tab[3])
s <- sqrt(sum(1/tab))
exp(log(or) + qnorm(c(.025, .975))*s)
```
