---
title: "Exercises 5"
author: "Jan Hohenheim"
output:
  pdf_document: default
  html_notebook: default
---

## Setup

```{r loading}
rm(list = ls())
library(ggplot2)
library(ggthemes)
library(dplyr)
library(purrr)
library(cowplot)
library(latex2exp)
library(glue)

theme <- 
  theme_solarized() #+ 
  #theme(text = element_text(family = "Helvetica Neue"))
```

```{r data}
HbSS <- c(7.2, 7.7, 8, 8.1, 8.3, 8.4, 8.4, 8.5, 8.6, 8.7, 9.1, 9.1, 9.1, 9.8, 10.1,
10.3)
HbSb <- c(8.1, 9.2, 10, 10.4, 10.6, 10.9, 11.1, 11.9, 12, 12.1)
Hb <- data.frame(level = c(HbSS, HbSb), 
                 category = c(rep("HbSS", length(HbSS)), rep("HbSb", length(HbSb))))
print("HbSS")
Hb |> 
  filter(category == "HbSS") |> 
  select(level) |>
  glimpse()

print("HbSb")
Hb |> 
  filter(category == "HbSb") |> 
  select(level) |>
  glimpse()
```

## Problem 17
Use again the sickle-cell disease data introduced in Problem 13 (Worksheet 4).
For the cases listed below, specify the null and alternative hypothesis. Then use R to perform the tests and give a careful interpretation.

### a) $\mu_{\rm{HbS\beta}} = 10$ ($\alpha = .05$, two-sided)
- $H_0: \mu_{\rm{HbS\beta}} = 10$
- $H_1: \mu_{\rm{HbS\beta}} \neq 10$
```{r p17a}
HbSb |>
  t.test(mu = 10, conf.level = 0.95)
```
10 is in the CI, accept $H_0$

### b) $\mu_{\rm{HbS\beta}} = \mu_{\rm{HbSS}}$ ($\alpha = .001$, two-sided)
- $H_0: \mu_{\rm{HbS\beta}} = \mu_{\rm{HbS}}$
- $H_1: \mu_{\rm{HbS\beta}} \neq \mu_{\rm{HbS}}$

```{r p17b}
HbSb |>
  t.test(HbSS, conf.level = 0.999)
```
0 is not in CI, reject $H_0$

### (c) What changes, if one-sided tests are performed instead?
The $H_0$ doesn't change, but the $H_1$ can be either:
- left-tailed: $H_1^{'}: \mu < n$ or
- right-tailed: $H_1^{''}: \mu > n$

```{r p17c}
HbSb |> t.test(mu = 10, conf.level = 0.95, alternative = "less")
HbSb |> t.test(mu = 10, conf.level = 0.95, alternative = "greater")
HbSb |> t.test(HbSS, conf.level = 0.999, alternative = "less")
HbSb |> t.test(HbSS, conf.level = 0.999, alternative = "greater")
```
All one-sided test variations accept $H_0$, except for the last, which leads us to accept the alternative $H_1: \mu_{\rm{HbS\beta}} > \mu_{\rm{HbSS}}$.
## Problem 18
Anorexia is an eating disorder that is characterized by low weight, food restriction, fear of gaining weight and a strong desire to be thin. The dataset *anorexia* in the package *MASS* gives the weight of 29 females before and after a cognitive behavioral treatment (in pounds). Test whether the treatment was effective.

- $H_0: \mu_{\rm{control}} = \mu_{\rm{cbt}}$
- $H_1: \mu_{\rm{control}} < \mu_{\rm{cbt}}$  
using $\alpha = 0.05$

```{r p18}
library(MASS)
?anorexia

anorexia.delta <- anorexia |>
  mutate(Deltawt = Postwt - Prewt)

control <- anorexia.delta |>
  filter(Treat == "Cont")
cbt <- anorexia.delta |>
  filter(Treat == "CBT")

alpha <- 0.05
t.test(control$Deltawt, cbt$Deltawt, alternative = "less", conf.level = 1 - alpha)
```
$p > 0.05$, so accept $H_0$: The CBT treatment had no significant effect.
