---
title: "exercises01"
author: "Jan Hohenheim"
date: "2/24/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 1 (Introduction to R/RStudio)
The aim of this exercise is to get some insight on the capabilities of the statistical software environment R and
the integrated development environment __RStudio__.

### (a) **R** has many built-in datasets, one example is __volcano__. Based on the help of the dataset, what is the name of the Volcano? Describe the dataset in a few words.
```{r volcano}
?volcano
```

The name of the volcano is Maunga Whau (Mt Eden). The data set is the height of the terrain in a given square in a grid, where each cell represents a terrain of 10 m by 10 m.

### (b) Use the **R** help function to get information on how to use the *image()* function for plotting matrices. Display the volcano data.
```{r volcano-image, echo=FALSE}
?image
image(volcano)
```
### (c) Install the package __fields__. Display the volcano data with the function __image.plot()__. What is the maximum height of the volcano?
```{r fields}
# install.packages("fields")
require(fields)

max(volcano)
image.plot(volcano)
```

The highest point is 195 m high.

### (d) Use the the R help function to find out the purpose of the function __demo()__ and have a look at the list of available demos. The demo of the function __persp()__ utilizes the volcano data to illustrate basic three-dimensional plotting. Call the demo and have a look at the plots.
```{r persp, echo=FALSE}
?demo
demo(persp)
```

## Problem 2 (EDA of bivariate data)
On www.isleroyalewolf.org/data/data/home.html the file isleroyale_graph_data_28Dec2011.xlsx contains popula-
tion data from wolves and moose. The information in this file is extracted and saved in the file __01wolvesmoose.csv__.
Download and read the file __01wolvesmoose.csv__ from the STA120 course page.

### (a) Construct a boxplot and a QQ-plot of the moose and wolf data. Give a short interpretation.
```{r qq}
par( mfcol=c(2, 2))
data <- read.csv("01_wolvesmoose.csv")
?boxplot
boxplot(data$Wolf, main="Wolf population")
boxplot(data$Moose, main="Moose population")

qqnorm(data$Wolf, main="Wolf population vs normal distribution")
qqline(data$Wolf, col=2, main='')  # add read line 
qqnorm(data$Moose, main="Moose population vs normal distribution")
qqline(data$Moose, col=2, main='')  # add read line 
```

The wolf and moose populations seem to follow a very similar distribution, with the moose population being skewed to the right. The normal distribution fits for all but the outer of the 5 quantiles.

### (b) Jointly visualize the wolves and moose data, as well as their abundances over the years. Give a short interpretation of what you see in the figures. (Of course you may compare the result with what is given on the aforementioned web page).

```{r joint}
pairs(x=data, main="Moose and wolf population against each other and years")
```

The populations seem to follow the Lotka-Volterra model: The wolf population lags behind the moose population, they both oscillate. Plotting the populations against each other reveals that a large amount of either population is not compatible with a large amount of the other.

## Problem 3 (EDA of multivariate data)
In this problem we want to explore the classical __mtcars__ dataset (directly available through the package __datasets__).
Perform an EDA thereof and provide at least three meaningful plots (as part of the EDA) and a short description
of what they display. In what measurement scale are the variables stored and what would be the natural or
original measurement scale?

```{r eda, echo=FALSE}
require(datasets)
?mtcars
str(mtcars)
summary(mtcars)
which(is.na(mtcars))
par(mfcol=c(1,2))
boxplot(mtcars[3:4])
boxplot(cbind(mtcars[1:2], mtcars[5:11]))
```
```{r eda3}
pairs(mtcars)
```

**Data collection**: Extracter from 1974 *Motor Trend* US magazine. No missing values.
Types of data:  
- **Discrete**: cyl (Number of cylinders), vs (Engine, 0 = V-shaped, 1 = straight), gear (Number of forward gears), carb (Number of carburetors), am (Transmission, 0 = automatic, 1 = manual)
- **Continuous**: mpg (Miles/(US) gallon), disp (Displacement in cubic inches), hp (Gross horsepower), drat (Rear axle ratio), wt (Weight in 1000s of lbs), qsec (1/4 mile speed)

```{r eda2}
plot(mtcars$wt, 
     mtcars$mpg, 
     main="Miles per gallon vs weight", 
     xlab="weight (1000s of lbs)", 
     ylab="Miles (US gallon)")
plot(mtcars$cyl, 
     mtcars$mpg, 
     main="Miles per gallon vs number of cylinders", 
     xlab="number of cylinders", 
     ylab="Miles (US gallon)")
plot(mtcars$disp, 
     mtcars$hp, 
     main="Displacement vs horsepower", 
     xlab="gross horsepower", 
     ylab="displacement (cubic inches)")
```

It seems that the miles per gallon correlates negatively with the weight and the number of cylinders. It also seems that the displacement correlates with the horsepower.

