---
title: "STA120 Exercises 3"
author: "Jan Hohenheim"
date: "3/7/2022"
output:
  html_notebook: default
  pdf_document: default
  html_document:
    df_print: paged
---

## Setup

```{r loading, echo = FALSE}
rm(list = ls())
library(ggplot2)
library(ggthemes)
library(dplyr)
library(purrr)
library(cowplot)
library(latex2exp)
library(glue)

theme <- 
  theme_solarized() + 
  theme(text = element_text(family = "Helvetica Neue"))
```


## Problem 8
In this problem we visualize and derive some properties of the Poisson random variable with parameter $\lambda > 0$

### a) 
Visualize in R the cdf and pmf of X ~ Pois($\lambda$), for $\lambda = 1$ and $\lambda = 5.5$
```{r p8a}

plot_lambda <- function(lambda) {
  n <- 15
  x <- 0:n 
  pmf <- x |>
    map_dbl(dpois, lambda)
  
  cdf <- x |>
    map_dbl(ppois, lambda)
  
  pmf_plot <- data.frame(x, probability = pmf) |>
    ggplot(aes(x = x, y = probability)) +
    geom_line(colour = "skyblue", alpha = 0.3, size = 1.2) +
    geom_point(colour = "salmon") +
    labs(title = TeX(glue("pmf of X ~ Pois($\\lambda$) with $\\lambda = {lambda}$")),
         x = "x",
         y = "f(x)") +
    theme
  
  cdf_plot <- data.frame(x, cummulative_probability = cdf) |>
    ggplot(aes(x = x, y = cummulative_probability)) +
    geom_line(colour = "skyblue", alpha = 0.3, size = 1.2) +
    geom_point(colour = "salmon") +
    labs(title = TeX(glue("cdf of X ~ Pois($\\lambda$) with $\\lambda = {lambda}$")), 
         x = "x",
         y = TeX("P(X $\\leq$ x)")) +
    theme
  
  plot_grid(pmf_plot, cdf_plot)
}
plot_lambda(1)
plot_lambda(5.5)
```

### b) 
For $\lambda = 1$ and $\lambda = 5.5$, sample from X1, ... , $X_n$ ~ Pois($\lambda$) with n = 200 and draw histograms. Compare the histograms with a). What do you expect to happen when n increases?

```{r p8b}
plot_pois_hist <- function(lambda){
  n <- 200
  plot <- data.frame(x = rpois(n, lambda)) |>
    ggplot(aes(x = x)) +
    geom_histogram(fill = "salmon", bins = 30, color = "white") +
    theme
  plot
}
plot_pois_hist(1)
plot_pois_hist(5.5)
```
These plots will approach the pmfs with higher n

### c) 
Let X ~ Pois($\lambda$), with $\lambda$ = 3: calculate P(X $\leq$ 2), P(X < 2) and P(X $\geq$ 3)
```{r p8c}
lambda = 3

p1 <- ppois(2, lambda)
glue("P(X <= 2) = {p1}")

p2 <- ppois(1, lambda)
glue("P(X < 2) = P(X <= 1) = {p2}")

p3 <- ppois(2, lambda, lower.tail = FALSE)
glue("P(X >= 3) = P(X > 2) = {p3}")
```

### d)
Plot the pmf of X ~ Pois($\lambda$), $\lambda$ = 5, and Y ∼ Bin(n, p) for n = 10, 100, 1000 with $\lambda$ = np. What can you conclude?

```{r p8d}
plot_pmfs <- function(n) {
  x <- 1:n
  lambda <- 5
  pois_pmf <- x |>
    map_dbl(dpois, lambda = lambda)
  p <- lambda / n
  bin_pmf <- x |>
    map_dbl(dbinom, size = n, prob = p)
  plot_pois <- data.frame(x = x, probability = pois_pmf) |>
    ggplot(aes(x = x, y = probability)) +
    geom_line(colour = "skyblue", alpha = 0.3, size = 1.2) +
    geom_point(color = "salmon") +
    labs(title = TeX(glue("X ~ Pois($\\lambda$) with $\\lambda = {lambda}$")),
         x = "x",
         y = "f(x)") +
    xlim(0, 13) +
    theme
  plot_binom <- data.frame(x = x, probability = bin_pmf) |>
    ggplot(aes(x = x, y = probability)) +
    geom_line(colour = "skyblue", alpha = 0.3, size = 1.2) +
    geom_point(color = "salmon") +
    xlim(0, 13) +
      labs(title = TeX(glue("X ~ Bin(n, p), n = {n}, p = {p}")),
         x = "x",
         y = "f(x)") +
    theme
  plot_grid(plot_pois, plot_binom)
}

ns <- c(10, 100, 1000)
lapply(ns, plot_pmfs)
```
The graphs are pretty much the same, the two distributions approach each other.

## Problem 10
Let X1, ... , $X_n$ be independent and identically distributed (iid) random variables following Exp($\lambda$). Assume $\lambda$ = 2
for the following.

### a)
Sample n = 100 random numbers from Exp($\lambda$). Visualize the data with a histogram and superimpose the theoretical density.
```{r q10}
n <- 100
lambda <- 2
sample = rexp(n, lambda)

x <- 0:(n - 1)
delta <- 0.2
theoretical <- x |>
  map_dbl(\(x) pexp(x + delta, lambda) - pexp(x, lambda)) * n


plot <-
  data_frame(x = x, sample = sample, theoretical = theoretical) |>
  ggplot() +
  geom_histogram(aes(x = sample), bins = 50, fill = "salmon", color = "white") +
  geom_point(aes(x = x, y = theoretical), colour = "skyblue") +
  geom_line(aes(x = x, y = theoretical), colour = "skyblue", size = 1.4, alpha = 0.4) +
  xlim(0, 15) +
  labs("X ~ Exp(2), sample and theory",
       x = "x",
       y = "count (real and theoretical") +
  theme
plot
```
### c)
Draw a histogram of min(X1, . . . , X100) from 500 realizations and compare it to the theoretical result from part b)
```{r p10c}
lambda <- 2
n <- 100
m <- 500
sample <- 1:m |>
  map_dbl(\(.) rexp(n, lambda) |> min())

plot <- 
  data.frame(x = sample) |>
  ggplot(aes(x = x)) +
  geom_histogram(fill = "salmon", color = "white", bins = 50) +
  theme
plot
```
It seems to resemble a normal distribution with a $\mu$ a bit bigger than 0
