---
title: "R Notebook"
author: "Jan Hohenheim"
output:
  pdf_document: default
  html_notebook: default
---

## Setup

```{r loading}
rm(list = ls())
library(ggplot2)
library(ggthemes)
library(dplyr)
library(purrr)
library(cowplot)
library(latex2exp)
library(glue)

theme <- 
  theme_solarized() #+ 
  #theme(text = element_text(family = "Helvetica Neue"))
```

## Problem  13
The following hemoglobin levels of blood samples from patients with HbSS and HbS/$\beta$ sickle cell disease are given (Statistische Prinzipien für medizinische Projekte. Hüsler, J. and Zimmermann, H. (2010)):

```{r p13}
HbSS <- c(7.2, 7.7, 8, 8.1, 8.3, 8.4, 8.4, 8.5, 8.6, 8.7, 9.1, 9.1, 9.1, 9.8, 10.1,
10.3)
HbSb <- c(8.1, 9.2, 10, 10.4, 10.6, 10.9, 11.1, 11.9, 12, 12.1)
Hb <- data.frame(level = c(HbSS, HbSb), 
                 category = c(rep("HbSS", length(HbSS)), rep("HbSb", length(HbSb))))
print("HbSS")
Hb |> 
  filter(category == "HbSS") |> 
  select(level) |>
  glimpse()

print("HbSb")
Hb |> 
  filter(category == "HbSb") |> 
  select(level) |>
  glimpse()
```
### (a) Visualize the data with boxplots.

```{r p13a}
plot <- Hb |>
  group_by(category) |>
  ggplot(aes(x = category, y = level)) +
  geom_boxplot() +
  theme
plot
```

### (b) Propose a statistical model for both diseases. What are the parameters? Estimate all parameters from your model based on intuitive estimators.

I propose:
- HbSb ~ N(10.75, 1.5) and 
- HbSS ~ Bin(8.5, 1)

### (c) Based on boxplots and QQ-plots, is there coherence between your model and the data?
```{r p13c}
Hb_sample <- Hb |>
  mutate(sample = c(rnorm(length(HbSS), mean = 8.5, sd = 1),
                    rnorm(length(HbSb), mean = 10.75, sd = 1.5)))
plot_box <- Hb_sample |>
  group_by(category) |>
  ggplot(aes(x = category, y = sample)) +
  geom_boxplot() +
  theme
plot_box

plot_qq <- Hb_sample |>
  ggplot(aes(sample = level)) +
  geom_qq() +
  geom_qq_line(colour = "maroon") +
  theme 
plot_qq
```
It fits pretty well!

## Problem 14
The dataset Oral is available in the R package spam and contains oral cavity cancer counts for 544 districts in Germany.

### (a) Load the data and take a look at its help page using ?Oral
```{r p14a}
library(spam)
?Oral
```

### (b) Compute summary statistics for all variables in the dataset. Which of the 544 regions has the highest number of expected counts E ?

```{r p14b}
Oral |> 
  glimpse()

Oral |>
  summary()

max_e_index <- Oral$E |>
  which.max()

print(glue("The region number {max_e_index} has the highest E at {max(Oral$E)}"))
```

### (c)
Poisson distribution is common for modeling rare events such as death caused by cavity cancer (column Y in the data). However, the districts differ greatly in their populations. Define a subset from the data, which only considers districts with expected fatal casualties caused by cavity cancer between 35 and 45 (subset, column E). Perform a Q-Q Plot for a Poisson distribution.  Hint: use qqplot() from the stats package and define the theoretical quantiles with qpois(ppoints(...), lambda=...)

```{r p14c}
Oral.subset <- Oral |>
  filter(35 >= E & E <= 45)

Oral.subset |> glimpse()

lambda <- round(mean(Oral.subset$Y))
print(glue("Using lambda = {lambda}"))
plot <- Oral.subset |>
  ggplot(aes(sample = E)) +
  geom_qq(distribution = qpois, dparams = lambda) +
  geom_qq_line(distribution = qpois, dparams = lambda, colour = "maroon") +
  theme
plot
```
### (d)
Simulate a Poisson distributed random variable with the same length and and the same lambda as your
subset. Perform a QQ-plot of your simulated data. What can you say about the distribution of your subset
of the cancer data?
```{r p14d}
sample <- rpois(nrow(Oral), lambda = lambda)
plot <- sample |>
  data.frame() |>
  ggplot(aes(sample = sample)) +
  geom_qq(distribution = qpois, dparams = lambda) +
  geom_qq_line(distribution = qpois, dparams = lambda, colour = "maroon") +
  theme
plot
```
> The assumed lambda does not fit very well

### (e)
Assume that the standardized mortality ratio Zi = Yi/Ei is normally distributed, i.e., Z1,..., Z544 iid ~ N ($\mu$, $\sigma^2$). Estimate $\mu$ and give a 95% (exact) confidence interval (CI). What is the precise meaning of the CI?
```{r p14e}
Z <- Oral$Y / Oral$E
mu <- mean(Z)
n <- length(Z)
sigma <- sd(Z)
alpha <- 0.05
ci <- mu + qt(p = c(alpha / 2, 1 - alpha / 2), df = n -  1) * sigma / sqrt(n)
print(glue("Estimating Z ~ N({mu}, {sigma^2})"))
print(glue("95% confidence interval: {ci[[1]]} - {ci[[2]]}"))
```
> The 95% CI means that when repeating the sampling many times, 95% of the time the mean will lie in the CI.

### (f)
Simulate a 95% confidence interval based on the following bootstrap scheme (sampling with replacement):
Repeat 10'000 times
– Sample 544 observations Zi with replacement
– Calculate and store the mean of these sampled observations
Construct the confidence interval by taking the 2.5% and the 97.5% quantiles of the stored means.
Compare it to the CI from e).
```{r p14e\f}
n <- 544
means <- 1:10000 |>
  map(\(.) rnorm(n, mu, sigma) |> mean()) |>
  unlist()
quantile(means, probs = c(0.025, 0.975))
```
> The data fits extremely well


